\vspace{1.0cm} \LARGE \bf Reassortment between influenza B lineages and the emergence of a co-adapted PB1-PB2-HA gene complex

### 2014-07-11 fluB.tex
Fix #37

Another possibility is the action of balancing selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{genomeGrid}, \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot}).
Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \citep{chen2006,obenauer2006}.
This is easiest to interpret as frequency-dependent balancing selection acting to preserve antigenic diversity \citep{worobey2014}.
If this is the case in influenza B viruses, we expect balancing selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.

@@

It also remains unknown whether reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background reassorting into a Victoria PB1-PB2-HA background are indicative of rare events followed by selective sweeps or stochastic fixation.
It's not unfeasible for selective sweeps following reassortments to break down developing co-adaptation of segments, especially if they are not functionally linked and have themselves been reassorted into a new PB1-PB2-HA background recently.
This is the second scenario we might expect to occur, whereby relatively frequent reassortments occur between Vic and Yam lineages and are followed by selective sweeps.
In this case influenza B viruses would undergo periodic genome homogenization events with the exception of PB1, PB2 and HA segments.
Because Vic HA and Yam HA are antigenically dissimilar we think that balancing selection would prevent even strong selective sweeps from driving the opposing PB1-PB2-HA gene complex to extinction.
This `\textit{status quo}' model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.

Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether dynamics similar to Victoria and Yamagata lineages have not occurred in influenza B virus genomes before and left no trace through extinction.
We find it unlikely that either Victoria or Yamagata lineage PB1-PB2-HA complexes will go extinct stochastically in the near future, as they have co-circulated for prolonged periods at a ratio close to 0.5, suggesting the action of balancing selection (Figure \ref{lineageRatiosOverTime}).
Extinction through depletion of susceptible individuals, such as following influenza pandemics or mass vaccination seem unlikely as well.
Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines, which are usually applied to specific subsets of the population, do not produce lifelong immunity.

>>>>

Another possibility is the action of frequency dependent selection in preserving the diversity in one segment, whilst the other segments hitchhike along.
A good candidate for this would be HA, as it is now the sole bearer of substantial antigenic diversity within the influenza B population, as the Vic lineage NA segment went extinct in 2002 (Figures \ref{genomeGrid}, \ref{tmrcaOT}, \ref{lineageRatiosOverTime} and \ref{railroadPlot}).
Previous research has found that avian influenza A virus HA and NA segments, which are the primary vehicles of antigenicity, exhibit vast diversity when compared to `internal segments' (including PB1 and PB2 segments), which show much more recent TMRCAs and less variability at the amino acid level \citep{chen2006,obenauer2006}.
This is easiest to interpret as frequency-dependent selection acting to preserve antigenic diversity \citep{worobey2014}.
If this is the case in influenza B viruses, we expect frequency-dependent selection to act on HA and indirectly, through some unknown association with HA, on PB1 and PB2 segments.

@@

It also remains unknown whether reassortment events of the past 20 years (Figure \ref{railroadPlot}) which frequently involved the NP, MP and NS segments from a Yamagata PB1-PB2-HA background reassorting into a Victoria PB1-PB2-HA background are indicative of rare events followed by selective sweeps or stochastic fixation.
It's not unfeasible for selective sweeps following reassortments to break down developing co-adaptation of segments, especially if they are not functionally linked and have themselves been reassorted into a new PB1-PB2-HA background recently.
This is the second scenario we might expect to occur, whereby relatively frequent reassortments occur between Vic and Yam lineages and are followed by selective sweeps.
In this case influenza B viruses would undergo periodic genome homogenization events with the exception of PB1, PB2 and HA segments.
Because Vic HA and Yam HA are antigenically dissimilar we think that frequency-dependent selection would prevent even strong selective sweeps from driving the opposing PB1-PB2-HA gene complex to extinction.
This `\textit{status quo}' model would require strong selective sweeps and/or relatively frequent reassortments, neither of which seem to be lacking in the influenza B virus population.

Given the relatively recent explosion of sequence data available for influenza B, it is difficult to say whether dynamics similar to Victoria and Yamagata lineages have not occurred in influenza B virus genomes before and left no trace through extinction.
We find it unlikely that either Victoria or Yamagata lineage PB1-PB2-HA complexes will go extinct stochastically in the near future, as they have co-circulated for prolonged periods at a ratio close to 0.5, suggesting the action of frequency dependent selection (Figure \ref{lineageRatiosOverTime}).
Extinction through depletion of susceptible individuals, such as following influenza pandemics or mass vaccination seem unlikely as well.
Both Victoria and Yamagata lineage PB1-PB2-HA complexes survived the admittedly mild influenza pandemic in 2009 and influenza vaccines, which are usually applied to specific subsets of the population, do not produce lifelong immunity.
